Constructing Sailing Match Race Schedules: Round Robin Pairing Lists

### 2015-04-02 paper.tex
some changes from me

\section{Introduction}
%\coffee{1}
This work describes a round robin competition format that arises from sailing, known as match racing. Competitors, i.e. {\em skippers}, compete against each other in a series of matches taking place in rounds, known as {\em flights}.  A match is composed of two skippers, with each skipper in a boat on his own. Skippers in a match set off together, one on the port side, the other starboard, and first home wins that match. ISAF World Sailing rules dictate what makes a legal schedule and what criteria to optimise. This is a rich set of rules, 13 in all, and as far as we know, all published schedules have been produced by hand. This is a daunting task. A close inspection of these schedules reveals that most are illegal, violating many of the match race rules, many schedules are missing and some that are legal are far from optimal.

This paper describes the scheduling problem. We present a constraint programming solution and document how this model was constructed incrementally. We believe this illustrates how adaptable constraint programming is, i.e. given a fairly robust model this can be incrementally developed to address all of the required criteria. Our schedules are then constructed in a sequence of stages, not dissimilar to the hybrid and decomposition based approaches in \cite{lombardi2012}. Some ISAF published schedules are presented along with some of our own, some of these being new.  These schedules are esentially blank forms, made available to umpires and competition managers. These forms are then filled in with names of actual competitors, i.e. these schedules are reusable.

@@

\section{Problem Definition: Round Robin Pairing Lists}\label{sec:definition}
The guidelines for running match racing events~\cite{isaf} places a number of criteria on how the competing skippers are scheduled into flights and matches, known as pairing lists. Several of these criteria are applicable when the number of skippers exceeds the number of available boats, in which case skippers have to change in and out of boats, and are allowed time in the schedule to check and fine-tune the boat they change into. Typically, matches set off at 5 minute intervals with each match taking about 20 minutes. Therefore, if we have 10 skippers and 10 boats we have 45 matches. This results in 9 flights each of 5 matches.  Each flight takes about 50 minutes, consequently eight or nine flights a day is a reasonable achievement for most umpires. Consequently, the number of boats and skippers involved is typically small.

The criteria for generated match racing schedules (pairing lists) are given in ISAF ``International Umpires' and Match Racing Manual'' \cite{isaf}, section
M.2 Recommended Criteria for Round Robin Pairing Lists, and are detailed below.

\begin{framed}
\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item *When skippers have an even number of matches, they have the same number of port and starboard assignments.
\item *When skippers have an odd number of matches, the first half of the skippers
    will have one more starboard assignment.
%\item When there is an even number of flights, each skipper has the same number
%    of port and starboard assignments.
%\item When there is an odd number of flights, the first half of the skippers
%    will have one more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of
    the next flight.
\item No skipper should have more than two consecutive port or starboard
    assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as
    equally as possible.
\item In flights with five or more matches, no skipper should be in the
    next-to-last match in a flight and then in the first match of the next
    flight.
\item If possible, a skipper should be starboard when meeting the nearest
    lowest ranked skipper (i.e.\ \#1 will be starboard against \#2, \#3 will be
    starboard against \#4).
\item Close-ranked skippers meet in the last flight.
\item Minimise the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}
Note that criteria 10, 11 and 12 only apply when there are fewer boats than
skippers.
\end{framed}

We have rephrased criteria 2 \& 3 (denoted *) to clarify an error in \cite{isaf}. Note that the order of matches within a flight is significant and is constrained, as is the order of flights within a schedule and the position of skippers within a match. This permits fair schedules that provide the opportunity for skippers to change boats, take fair advantage of wind direction, etc.

Criteria 12 defines what is meant by a boat change: if in flight $i$ this skipper is not in a match (i.e. it is a bye for this skipper) but is in a match in flight $i+1$ then he has to get into a boat, rearrange all his kit and set the boat as he likes it prior to casting off, and this is a boat change.

Next, we note that skippers can be ranked prior to an event -- based on their performance in previous events~\footnote{\url{http://www.sailing.org/rankings/match/}} -- which is used to seed the skippers. The ordering of skippers within a match signifies their starting position (port or starboard) -- as skippers allocated the starboard starting position gain a small competitive advantage in that match - criteria 8 accomplishes the seedings.

@@

There are many redundant variables in our model. One of our first valid models was essentially that shown in Figure \ref{skipper5}, and did not use $modMatch$ and $match$ variables. Performance was hopeless, struggling to produce a 7 skipper 6 boat schedule in days. The $modMatch$ and $match$ variables were added. This improved domain filtering, and a flattened $match$ array was used as decision variables. That is, the decisions where "when do we schedule this match?".  At this point we had not yet produced a 9 skipper 6 boat schedule and we suspected that the combined criteria might exclude such a schedule. A non-constraint model was developed by the fourth author, primarily to prove that there was no schedule for 9 skippers with 6 boats. This program used a backtracking search with decision variables being positions within flights, i.e. time slots being assigned matches. Prior to search, symmetry breaking was performed, and at the top of search the matches in the first flight were set. A solution was found in seconds! With this experience we added in the $time$ variables, channeled these into the existing model and used these as decision variables, i.e. the question now was "what match will we do now?". We also anchored the matches in the first flight. With these changes we began to produce schedules in acceptable time. The model was then built upon, incrementally adding criteria. This took a surprisingly short amount of time, sometimes minutes of coding to add in a new feature. The model was then enhanced so that it would address optimisation rather than just satisfaction, again a trivial programming task.

@@

\section{Conclusion}\label{sec:conclusions}
We have produce new and better match race schedules. These schedules can be used by anyone that competes under ISAF World Sailing rules. Our schedules can be downloaded as blank schedules and then populated with the names of the actual competing skippers.

So, why did we use constraint programming? The answer is obvious:  we are constraint programmers, or to borrow Mark Twain's words "To a man with a hammer, everything looks like a nail." But actually, constraint programming has been a good way to go. From an engineering perspective, it has allowed us to prototype solutions quickly and to build solutions incrementally. There is also an advantage that we might exploit in the future: we can now investigate the effect different criteria have on our ability to produce schedules and how relaxation of those affect optimisation criteria. That is, the constraint program might be used to design the next set of umpire's rules.

>>>>

\section{Introduction}
%\coffee{1}
This work describes a round robin competition format that arises from sailing, known as match racing. Competitors, i.e. {\em skippers}, compete against each other in a series of matches taking place in rounds, known as {\em flights}.  A match is composed of two skippers, with each skipper in a boat on his own. Skippers in a match set off together, one on the port side, the other starboard, and first home wins that match. Criteria published by the Interational Sailing Federation (ISAF) dictate what makes a legal match race schedule and what should be optimised. This is a rich set of cirteria, 13 in all, and as far as we know, all published schedules have been produced by hand. This is a daunting task. A close inspection of these schedules reveals that most are illegal, violating many of the match race scheduling criteria, many schedules are missing and some that are legal are far from optimal.

This paper describes the scheduling problem. We present a constraint programming solution and document how this model was constructed incrementally. We believe this illustrates how adaptable constraint programming is, i.e. given a fairly robust model this can be incrementally developed to address all of the required criteria. Our schedules are then constructed in a sequence of stages, not dissimilar to the hybrid and decomposition based approaches in \cite{lombardi2012}. Some ISAF published schedules are presented along with some of our own, some of these being new.  These schedules are esentially blank forms, made available to event organisers, which can be filled in with names of actual competitors, i.e.\ these schedules are reusable.

@@

\section{Problem Definition: Round Robin Pairing Lists}\label{sec:definition}
The guidelines for running match racing events~\cite{isaf} places a number of criteria on how the competing skippers are scheduled into flights and matches, known as pairing lists. Several of these criteria are applicable when the number of skippers exceeds the number of available boats, in which case skippers have to change in and out of boats, and are allowed time in the schedule to check and fine-tune the boat they change into. Typically, matches set off at 5 minute intervals with each match taking about 20 minutes. Therefore, if we have 10 skippers and 10 boats we have 45 matches. This results in 9 flights each of 5 matches.  Each flight takes about 50 minutes, consequently eight or nine flights a day is a reasonable achievement for most events~\cite{isaf}. Consequently, the number of boats and skippers involved is typically small.

The criteria for  match racing schedules (pairing lists) are given in ISAF ``International Umpires' and Match Racing Manual'' \cite{isaf}, section
M.2 Recommended Criteria for Round Robin Pairing Lists, and are detailed below.

\begin{framed}
Principal Criteria in Order of Priority:
\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item *When skippers have an even number of matches, they have the same number of port and starboard assignments.
\item *When skippers have an odd number of matches, the first half of the skippers
    will have one more starboard assignment.
%\item When there is an even number of flights, each skipper has the same number
%    of port and starboard assignments.
%\item When there is an odd number of flights, the first half of the skippers
%    will have one more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of
    the next flight.
\item No skipper should have more than two consecutive port or starboard
    assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as
    equally as possible.
\item In flights with five or more matches, no skipper should be in the
    next-to-last match in a flight and then in the first match of the next
    flight.
\item If possible, a skipper should be starboard when meeting the nearest
    lowest ranked skipper (i.e.\ \#1 will be starboard against \#2, \#3 will be
    starboard against \#4).
\item Close-ranked skippers meet in the last flight.
\item Minimise the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}
Note that criteria 10, 11 and 12 only apply when there are fewer boats than
skippers; 11 and 12 override 6 when changes are required.
\end{framed}

We have rephrased criteria 2 \& 3 (denoted *) to clarify an error in \cite{isaf}. Note that the order of matches within a flight is significant and is constrained, as is the order of flights within a schedule and the position of skippers within a match. This permits fair schedules that provide the sufficient changeover time for skippers changing boats, etc.

Criteria 12 defines what is meant by a boat change: if in flight $i$ this skipper is not in a match (i.e. it is a bye for this skipper) but is in a match in flight $i+1$ then he has to get into a boat, rearrange all his kit and set the boat to his preference before racing commences, and this is a boat change.

Next, we note that skippers can be ranked prior to an event -- based on their performance in previous events\footnote{\url{http://www.sailing.org/rankings/match/}} -- which is used to seed the skippers. The ordering of skippers within a match signifies their starting position (port or starboard) -- as skippers allocated the starboard starting position gain a small competitive advantage in that match - criteria 8 accomplishes the seedings.

@@

There are many redundant variables in our model. One of our first valid models was essentially that shown in Figure \ref{skipper5}, and did not use $modMatch$ and $match$ variables. Performance was hopeless, struggling to produce a 7 skipper 6 boat schedule in days. The $modMatch$ and $match$ variables were added. This improved domain filtering, and a flattened $match$ array was used as decision variables. That is, the decisions where ``when do we schedule this match?".  At this point we had not yet produced a 9 skipper 6 boat schedule and we suspected that the combined criteria might exclude such a schedule. A non-constraint model was developed by the fourth author, primarily to prove that there was no schedule for 9 skippers with 6 boats. This program used a backtracking search with decision variables being positions within flights, i.e. time slots being assigned matches. Prior to search, symmetry breaking was performed, and at the top of search the matches in the first flight were set. A solution was found in seconds! With this experience we added in the $time$ variables, channeled these into the existing model and used these as decision variables, i.e. the question now was ``what match will we do now?". We also anchored the matches in the first flight. With these changes we began to produce schedules in acceptable time. The model was then built upon, incrementally adding criteria. This took a surprisingly short amount of time, sometimes minutes of coding to add in a new feature. The model was then enhanced so that it would address optimisation rather than just satisfaction, again a trivial programming task.

@@

\section{Conclusion}\label{sec:conclusions}
We have produce new and better match race schedules. These schedules can be used by anyone that competes under the criteria published by ISAF. Our schedules can be downloaded as blank schedules and then populated with the names of the actual competing skippers.

So, why did we use constraint programming? The answer is obvious:  we are constraint programmers, or to borrow Mark Twain's words ``To a man with a hammer, everything looks like a nail." But actually, constraint programming has been a good way to go. From an engineering perspective, it has allowed us to prototype solutions quickly and to build solutions incrementally. There is also an advantage that we might exploit in the future: we can now investigate the effect different criteria have on our ability to produce schedules and how relaxation of those affect optimisation criteria. That is, the constraint program might be used to design the next set of criteria for match race schedules, or allow the event organiser to decide which criteria he may sacrifice to ensure a faster schedule with less boat changes.
